\documentclass[output=paper,
modfonts
]{LSP/langsci}


%\input{localpackages.tex}
%\input{localcommands.tex}

\title{\emph{u}-umlaut in Icelandic and Faroese: Survival and death}

\author{%
Höskuldur Thráinsson\affiliation{University of Iceland}
}

\abstract{Although Icelandic and Faroese are closely related and very similar in
many respects, their vowel systems are quite different (see e.g.
\citealt{anderson1969a,arnason2011}). This paper compares \emph{u}-umlaut alternations
in Icelandic and Faroese and shows that the Faroese umlaut has a number
of properties that are to be expected if the relevant alternations are
morphological (or analogical) rather than being due to a synchronic
phonological process. In Icelandic, on the other hand, \emph{u}-umlaut has none
of these properties and arguably behaves like a living phonological
process. This is theoretically interesting because the quality of the
vowels involved (both the umlaut trigger and the target) has changed
from Old to Modern Icelandic. In addition, \emph{u}-umlaut in Modern Icelandic
is more opaque (in the sense of \citealt{kiparsky1973}) than its Old Icelandic
counterpart, i.e.\ it has more surface exceptions. An epenthesis rule
inserting a (non-umlauting) /u/ into certain inflectional endings is the
cause of many of these surface exceptions. Yet it seems that \emph{u}-umlaut in
Icelandic is still transparent enough to be acquired by children as a
phonological process. In Faroese, on the other hand, \emph{u}-umlaut became too
opaque and died out as a phonological rule. It is argued that this has
partly to do with certain changes in the Faroese vowel system and partly
with the fact that the u-epenthesis rule was lost in Faroese.}

%\protect\hypertarget{__RefHeading__452_2075933062}{}{}\emph{\textbf{\\
%}} % commented out because it prevented compilation

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

Anderson put the process of \emph{u-}umlaut in Icelandic on the modern linguistic map with the analysis he proposed in his dissertation \citep{Anderson1969} and several subsequent publications \citep{anderson1969b,anderson1972t,anderson1973,anderson1974,anderson1976t}. Because of changes in the vowel system from Old to Modern
Icelandic, the nature of the umlaut process changed somewhat through the ages (see e.g. \citealt{benediktsson1959}). The most important part of \emph{u-}umlaut, and the only part that is alive in the modern language, involves /a/ \textasciitilde{} /ǫ/ alternations in the old language (phonetically {[}a{]} \textasciitilde{} {[}ɔ{]}, as shown in (2)), which show up as /a/ \textasciitilde{} /ö/ alternations in the modern language (phonetically {[}a{]} \textasciitilde{} {[}œ{]}, cf. (2)). This is illustrated in (1) with the relevant vowel symbols highlighted:

\ea
\begin{tabular}[t]{ l l }
Old Icelandic: & Modern Icelandic: \\
\emph{saga} ʽsagaʼ, \textsc{obl} \emph{sǫgu}, \textsc{pl} \emph{sǫgur} & \emph{saga}, \textsc{obl} \emph{sögu}, \textsc{pl} \emph{sögur}\\

\emph{hvass} `sharp', \textsc{dat} \emph{hvǫssum} & \emph{hvass}, \textsc{dat} \emph{hvössum}\\

\emph{tala} `speak', \textsc{1.pl} \emph{tǫlum} & \emph{tala}, \textsc{1pl} \emph{tölum}\\
\end{tabular}
\z

\noindent As these examples suggest, the quality of the root vowel /a/ changes when a /u/ follows in the next syllable. The relevant proecesses can be illustrated schematically as in (2). For the sake of simplicity I use conventional orthographic symbols to represent the vowels and only give IPA-symbols for the vowels that are important for the understanding of the umlaut processes. The umlaut-triggering vowels are encircled:\footnote{Note that in the representation of the Modern Icelandic vowel system, the accents over vowel symbols have nothing to do with quantity but simply denote separate vowel
qualities. Thus /í/ is {[}i{]}, /i/ is {[}ɪ{]}, /ú/ is {[}u{]} and /u/ is {[}ʏ{]}, as the schematic representation in (2) suggests.}

\ea
	\ea \emph{u-}umlaut in Old Icelandic and the system of short vowels:\\
	\begin{tabular}{ccccccc}
           		 && \multicolumn{2}{c}{{[}-back{]}} &  & \multicolumn{2}{c}{{[}+back{]}} \\
           		 && {[}-round{]}   	& {[}+round{]}   &  & {[}-round{]}   & {[}+round{]}   \\
	       {[}+high{]}    & &      i         &    y            &  &                &    \textcircled{u} {[}u{]}            \\
           			 & &      e       &    ø          &  &                &     o           \\
	       {[}+low{]}  &	  &    ę        &                 &  &          a \tikzmarkfullnamed{a}{[a]}   & \tikzmarkfullnamed{o}{ǫ} {[}ɔ{]} \\           
	\end{tabular}\tikzarrow{a}{o}
%\newpage
	\ex \emph{u-}umlaut in Modern Icelandic and the system of monophthongs:\footnote{I am 			assuming here, like Thráinsson (1994) and Gíslason and Thráinsson (2000, p. 34), for 			instance, that Modern Icelandic only distinguishes between three three vowel heights and that /e/ 	{[}ɛ{]} and /ö/ {[}œ{]} are both phonologically {[}+low{]}, like /a/ {[}a{]} and /o/ {[}ɔ{]}. For different 	assumptions see e.g. \citet[60]{arnason2011}.}\\
	\begin{tabular}{ccccccc}
           		 && \multicolumn{2}{c}{{[}-back{]}} &  & \multicolumn{2}{c}{{[}+back{]}} \\
           		 && {[}-round{]}   	& {[}+round{]}   &  & {[}-round{]}   & {[}+round{]}   \\
	       {[}+high{]}  &   &  í        &                &  &                &    ú {[}u{]}            \\
           			&  &   i    &   \textcircled{u} {[}ʏ{]} &  &                &                \\
	       {[}+low{]}  	&  &  e    &    ö \tikzmarkfullnamed{oe}{[œ]}           &  & \tikzmarkfull{a} {[}a{]}	    &          o {[}ɔ{]} \\
	\end{tabular}\tikzarrow{a}{oe}
	\z
\z

The gist of Anderson's analysis of \emph{u-}umlaut can then be illustrated semi-formally as in the traditional generative phonological notation in (3), with the assimilating features highlighted (see also
\citealt[31]{roegnvaldsson1981}, \citealt[89--90]{thrainsson2011}):\footnote{Here, and elsewhere in this paper, I will use the kinds of formulations of rules and conditions familiar from classical generative phonology since much of the work on \emph{u-}umlaut has been done in that kind of framework. For analyses employing more recent frameworks see \citealt{gibson2000}, \citealt{hansson2013} and \citealt{ingason2016}. Most of the argumentation in this paper should be relatively framework-independent, however.}

\ea
	\ea \emph{u-}umlaut in Old Icelandic:\\
	\phonr{/a/}{\phonfeat{\feat{\textbf{+round}}}}{C\sub{0}V\sub{\normalsize\phonfeat{\feat{\textbf{+round}}\\\feat{+high}\\\feat{+back}}}}	
	\ex \emph{u-}umlaut in Modern Icelandic:\\
	\phonr{/a/}{\phonfeat{\feat{\textbf{+round}}\\\feat{\textbf{-back}}}}{C\sub{0}V\sub{\normalsize\phonfeat{\feat{\textbf{+round}}\\\feat{\textbf{-back}}\\\feat{-low}}}}
	\z
\z

\noindent As the illustration in (3) shows, the modern version of the umlaut is
somewhat more complex than the old one, assimilating two features rather
than one. Nevertheless, it is still arguably a phonologically (or
phonetically) natural assimilation process, assimilating rounding and
backness.

Although the \emph{u-}umlaut discussion was most lively on the
international scene in the 1970s (see e.g. \citealt{iverson1978t,iverson1976,oresnik1975,oresnik1977}, cf. also \citealt{valfells1967}), the topic
keeps popping up to this day, e.g. in journals and conferences dedicated
to Scandinavian linguistics (see e.g. \citealt{gibson2000,indridason2010,thrainsson2011,hansson2013}) and even in recent master's theses
and doctoral dissertations (see \citealt{markusson2012,ingason2016}). The main
reason is that while \emph{u-}umlaut in Modern Icelandic is obviously
very productive, being applied consistently to new words and loanwords,
it shows a number of intriguing surface exceptions. These have been
discussed extensively in the literature cited but here I will
concentrate on the most common and widespread one, namely the lack of
umlaut before a /u/ that has been inserted between the inflectional
ending /r/ and a preceding consonant. This epenthesis did not exist in
Old Icelandic as illustrated in (4):

\ea
\begin{tabular}[t]{ l l }
Old Icelandic: & Modern Icelandic: \\
\emph{dalr} `valley', \emph{latr} `lazy'  & \emph{dal\textbf{u}r}, \emph{lat\textbf{u}r}
\end{tabular}
\z

\noindent If \emph{u-}umlaut is a phonological rule in the modern language, this
\emph{u-}epenthesis has to follow it, as it did historically. This is
one of the properties of \emph{u-}umlaut that have been used to argue
for the necessity of relatively abstract phonological representations
and derivations (e.g. \citealt{anderson1969a,anderson1974,roegnvaldsson1981,thrainsson2011,hansson2013}) while others have maintained that
\emph{u-}umlaut is not a phonological process anymore in Modern
Icelandic and the relevant alternations are morphologized and purely
analogical (see e.g. \citealt{markusson2012}) or at least ``morpheme-specific'',
i.e. triggered by particular morphemes that may or may not contain a /u/
(\citealt{ingason2016}).\footnote{\citet[220]{ingason2016} formulates his umlaut rule as follows:

  \begin{quote}
  Realize an underlying /a/ as /ö/ in the syllable which precedes the
  morpheme which triggers the umlaut.
  \end{quote}

  \noindent As can be seen here, no mention is made of a triggering /u/ in the
  rule. The reason is that Ingason wants to derive all all paradigmatic
  /a/ \textasciitilde{} /ö/ alternations the same way, including the
  ones where /u/ has been syncopated historically. Thus he argues that
  the \textsc{nom.sg.} morpheme -\textsc{ø} in feminine nouns like
  \emph{sök} `guilt, case' and the \textsc{nom./acc.pl.} morpheme
  -\textsc{ø} in neuter nouns like \emph{börn} `children' triggers
  umlaut the same way that the \textsc{dat.pl.} morpheme \emph{-um} does
  in \emph{sökum} and \emph{börnum}. But many researchers have wanted to
  distinguish between morphologically conditioned umlaut, where there is
  no triggering /u/, and phonologically conditioned umlaut triggered by
  /u/, e.g. \citet{roegnvaldsson1981}. One reason for doing so comes from the
  behavior of loanwords like the adjective \emph{smart} `smart, chic'.
  Here the \textsc{nom.sg.f} and the \textsc{nom/acc.pl.n} can either be
  \emph{smart} or \emph{smört}, i.e. a morphologically conditioned
  umlaut may or may not apply. But once an umlauting inflectional ending
  containing /u/ is added to the loanword \emph{smart}, the
  \emph{u-}umlaut becomes obligatory. Thus \textsc{dat.pl} can only be
  \emph{sm\textbf{ö}rt-um} and not *\emph{sm\textbf{a}rt-\textbf{u}m}
  and the \textsc{nom.pl.wk} form has to be
  \emph{sm\textbf{ö}rt-\textbf{u}} and not *\emph{smart-\textbf{u}}.
  This suggests that the morphologically conditioned umlaut is more
  prone to exceptions than the phonologically conditioned one, which is
  actually to be expected. Thanks to Eiríkur Rögnvaldsson for pointing
  this out to me.}

In this paper I will compare \emph{u-}umlaut alternations in Modern
Icelandic and Modern Faroese. This comparison will show very clearly
that \emph{u-}umlaut in Modern Faroese has a number of properties (e.g.
paradigm levelling, various kinds of exceptions, total absence from
certain paradigms, inapplicability to loanwords ...) that are to be
expected if the relevant alternations are no longer due to a synchronic
process. In Modern Icelandic, on the other hand, \emph{u-}umlaut has
none of these properties and behaves more like a phonological rule. This
is of general theoretical interest since it illustrates how phonological
rules can survive (in the case of Icelandic) despite reduced
transparency (in the sense of \citealt{kiparsky1973}) and how changes in the
phonological system can cause the death of a phonological rule (in the
case of Faroese) and what the consequences can be.

The remainder of the paper is organized as follows: In section 2 I first
illustrate how the \emph{u-}epenthesis works in Modern Icelandic and
then present a couple of arguments for the phonological (as opposed to
morphological) nature of Modern Icelandic \emph{u-}umlaut. Section 3
first describes some facts about the Faroese vowel system that must have
been important for the development of \emph{u-}umlaut and then shows
that \emph{u-}epenthesis does not exist anymore as a phonological
process in Modern Faroese. It is then argued that these developments led
to the death of \emph{u-}umlaut as a phonological process in Faroese.
Section 4 then contains a systematic comparison of \emph{u-}umlaut
alternations in Modern Icelandic and Faroese, concluding that the
Faroese ones must be analogical (and morphological) in nature as they do
not exhibit any of the crucial phonological properties that Modern
Icelandic \emph{u-}umlaut alternations show. In Icelandic, on the other
hand, \emph{u-}umlaut does not show the non-phonological properties
listed for its Faroese counterpart. Section 5 concludes the paper.

\section{\emph{u}-epenthesis and \emph{u-}umlaut in Modern Icelandic}

\subsection{The epenthesis rule}\label{the-epenthesis-rule}

The phoneme /r/ frequently occurs in Old Icelandic (Old Norse) as a
marker of various morphological categories, including \textsc{nom.sg} of
strong masculine nouns and adjectives as illustrated in (5). It
sometimes assimilated to a preceding consonant, e.g. /s, l, n/ (cf.
(5c)),\footnote{Assimilation to stem-final /l, n/ only happened in Old
  Icelandic if these consonants were preceded by long vowels, i.e. Old
  Icelandic diphthongs and vowels that are standardly represented by
  accented vowel symbols in Old Icelandic orthography, cf.
  \emph{stól-\textbf{l}} `chair' vs. \emph{dal-r} `valley',
  \emph{fín-\textbf{n}} `fine' vs. \emph{lin-r} `soft, limp',
  \emph{heil-\textbf{l}} `whole' vs. \emph{hol-r} `hollow'.} but it was
deleted after certain consonant clusters, such as /gl, gn, ss/ (cf.
(5d)):

\ea
	\ea \emph{stór-r} `big', \emph{mó-r} `peat', \emph{há-r} `high'
	\ex \emph{dal-r} `valley', \emph{lat-r} `lazy', \emph{tóm-r} `empty', \emph{harð-r} `hard'
	\ex \emph{ís-s} `ice', \emph{laus-s} `loose', \emph{stól-l} `chair', \emph{fín-n} `fine'
	\ex \emph{fugl} `bird', \emph{vagn} `wagon', \emph{foss} `waterfall' (stem \emph{foss-})
	\z
\z

\noindent It is likely that the /r/ in words of type (5b) was syllabic in Old
Icelandic. There are no syllabic consonants in Modern Icelandic, on the
other hand. Instead a /u/ appears between the /r/ and the preceding
consonant in the modern version of words of type (5b). There is
historical evidence for this \emph{u-}insertion from the late
thirteenth century and onwards (see e.g. \citealt{kristinsson1992}
and references cited there) and many linguists have argued that
\emph{u-}epenthesis still a productive phonological process in Modern
Icelandic (e.g.  \citealt{anderson1969a,anderson1969b,oresnik1972,roegnvaldsson1981,kiparsky1984}).\footnote{Orešnik later  maintained that
  \emph{u-}epenthesis could not be a synchronic rule in Modern Icelandic
  because of the existence of exceptional word forms like \emph{klifr}
  `climbing' (from the verb \emph{klifra} `climb'), \emph{sötr}
  `slurping' (from the verb \emph{sötra} `slurp'), \emph{pukr}
  `secretiveness' from the verb \emph{pukra(st)} `be secretive about',
  etc. (\citealt{oresnik1978}; see also the discussion in \citealt{kjartansson1984}). In words of this
  kind one would have expected \emph{u-}epenthesis to apply. The
  importance of these exceptions is not very clear since this is a very
  special class of words (all derived from verbs ending in -\emph{ra})
  and it is typically possible or even preferred to apply the epenthesis
  rule to these forms, giving \emph{klif\textbf{u}r},
  \emph{söt\textbf{u}r}, \emph{puk\textbf{u}r}, etc. For the sake of
  completeness it should be noted that the final \emph{-r} in word forms
  like \emph{sötr}, \emph{pukr} has to be voiceless and this may be
  related to the fact that there are no syllabic consonants in Modern
  Icelandic, as stated above.} This implies that speakers distinguish
between a \emph{-ur}-ending where the underlying morpheme is \#-r\# and
the /u/ is epenthetic (and does not trigger \emph{u-}umlaut) and a
-\emph{ur-}ending where the /u/ is not epenthetic and the underlying
morpheme is \#-ur\# (and the /u/ triggers \emph{u-}umlaut). This
contrast is illustrated in (6a) vs.\ (6b) (see also the examples in (1)
and (4) above):

\ea\begin{tabular}[t]{ll@{ }l@{ }l@{ \ }l@{ \ }l@{ }l@{ \ }l@{}}
a. & \#dal+r\# & `valley' & \textsc{nom.sg.m} & → & \emph{d\textbf{a}l-\textbf{u}r} & by epenthesis & no umlaut\\
&\#lat+r\# &`lazy' &\textsc{nom.sg.m} &→& \emph{l\textbf{a}t-\textbf{u}r} &by epenthesis &no umlaut\\
b. &\#sag+ur\#& `sagas' &\textsc{nom.pl.f}& →& \emph{s\textbf{ö}g-\textbf{u}r}& &\emph{u-}umlaut\\
&\#tal+ur\# &`numbers' &\textsc{nom.pl.f}& →& \emph{t\textbf{ö}l-\textbf{u}r}&& \emph{u-}umlaut\\
\end{tabular}
%	\ea \#dal+r\# `valley' \textsc{nom.sg.m} → \emph{d\textbf{a}l-\textbf{u}r} by epenthesis no umlaut\\
%	 \#lat+r\# `lazy' \textsc{nom.sg.m} → \emph{l\textbf{a}t-\textbf{u}r} by epenthesis no umlaut
%	 \ex \#sag+ur\# `sagas' \textsc{nom.pl.f} → \emph{s\textbf{ö}g-\textbf{u}r} \emph{u-}umlaut\\
%	\#tal+ur\# `numbers' \textsc{nom.pl.f} → \emph{t\textbf{ö}l-\textbf{u}r} \emph{u-}umlaut
%	\z
\z

\noindent Thus the \textsc{nom.sg} ending \#-r\#, which is both found in strong
masculine nouns like \emph{dalur} `valley' and in the strong masculine
form of adjectives like \emph{latur} `lazy', does not have the same
properties as the \textsc{nom.pl} ending \#-ur\# which is found in
feminine nouns like \emph{sögur} `sagas' and \emph{tölur} `numbers'.
Despite their surface similarities in certain environments, speakers can
clearly distinguish these endings. A part of the reason must be that the
\textsc{nom.sg.m} ending \#-r\# only shows up as -\emph{ur} in
phonologically definable environments, i.e. the modern version of words
with stems of type (5b), whereas the \textsc{nom.pl.f} ending \#-ur\# is
not so restricted and always shows up as -\emph{ur}. This is illustrated
in Table~\ref{tab:-r-ur} (compare the examples in (5)).

\begin{table}
\begin{tabularx}{\textwidth}{@{}l>{\raggedright}p{1in}>{\raggedright}X>{\raggedright}X@{}}
\lsptoprule
& type of stem & phonological realization of the \textsc{nom.sg.m} ending \#-r\# & phonological realization of the \textsc{nom.pl.f} ending \#-ur\#\tabularnewline
\midrule
a. & ending in a vowel & -\emph{r} \newline (\emph{mó-r} `peat', \emph{há-r}
`high') & \emph{\textbf{-ur}} \newline (\emph{ló-ur} `golden
plovers')\tabularnewline
b. & ending in a sin- gle consonant (but see c) & \textbf{-\emph{ur}} \newline (\emph{dal-ur} `valley', \emph{lat-ur} ʽlazyʼ) & \textbf{-\emph{ur}}
\newline (\emph{sög-ur} ʽsagasʼ, \emph{töl-ur} ʽnumbersʼ)\tabularnewline
c. & ending in a high vowel + /l,n/ & assimilation \newline (\emph{stól-l} `chair',
\emph{fín-n} `fine') & \textbf{-\emph{ur}} \newline (\emph{spús-ur} `wives',
\emph{súl-ur} ʽcolumnsʼ, \emph{dýn-ur} ʽmattresses')\tabularnewline
d. & ending in /s, r/ or consonant clusters ending in /l, n/ such as /gl,
gn/ & deletion \newline (\emph{ís} `ice', \emph{laus} `loose', \emph{foss}
`waterfall'\emph{bjór} `beer', \emph{stór} `big', \emph{fugl} `bird',
\emph{vagn} `wagon') & \textbf{-\emph{ur}} \newline (\emph{ýs-ur} ʽhaddocksʼ,
\emph{aus-ur} `scoops', \emph{hór-ur} `whores', \emph{ugl-ur} `owls',
\emph{hrygn-ur} ʽspawning fish', \emph{byss-ur} ʽgunsʼ)\tabularnewline
\lspbottomrule
\end{tabularx}
\caption{Phonological realization of the inflectional endings
	\#-r\# and \#-ur\# in Modern Icelandic.}
\label{tab:-r-ur}
\end{table}

Comparison of Table~\ref{tab:-r-ur} and the Old Icelandic examples in (5) reveals a
slight extension of \emph{r-}deletion: The /r/ of the morphological
ending \#-r\# is now deleted after /r/ (compare line d of the table to
(5a)) and after all instances of /s/, not just /ss/ (compare line d of
the table to (5c,d)). The \emph{u-}epenthesis illustrated in line b of
Table~\ref{tab:-r-ur} is an innovation, of course. Otherwise the \textsc{nom.sg.m}
ending behaves in much the same way as in Old Icelandic. The different
behavior of the morphemes compared in Table~\ref{tab:-r-ur} can be seen as an argument
for distinguishing them in the underlying form, e.g. for not analyzing
the \textsc{n.sg.m} ending as \#-ur\#.

\subsection{Some phonological properties of Modern Icelandic \emph{u}-umlaut}

In this section I will mention two sets of facts which show that
\emph{u-}umlaut still has certain properties in Modern Icelandic that
are to be expected if it is a phonologically conditioned process.

First, if \emph{u-}umlaut was morphologically conditioned and not
phonologically, we would expect it to be restricted to certain
morphological categories or parts of speech. It is not. It applies in
the paradigms of nouns, adjectives and verbs when a /u/ follows in the
inflectional ending (with the exception of the epenthetic /u/ already
mentioned). This is illustrated in (7):

\ea
	\ea	\emph{saga} `saga', \textsc{obl.sg}
		\emph{s\textbf{ö}g-\textbf{u}}, \textsc{nom/acc.pl}
		\emph{s\textbf{ö}g-\textbf{u}r}, \textsc{dat.pl}
		\emph{s\textbf{ö}g-\textbf{u}m}
	\ex	\emph{snjall} `smart', \textsc{dat.sg.m}
		\emph{snj\textbf{ö}ll-\textbf{u}m}, \textsc{nom.pl.wk}
		\emph{snj\textbf{ö}ll-\textbf{u}}
	\ex	\emph{kalla} `call', \textsc{1.pl.prs}
		\emph{k\textbf{ö}ll-\textbf{u}m}, \textsc{3.pl.pst}
		\emph{k\textbf{ö}ll\textbf{u}ð-\textbf{u}}
	\z
\z

\noindent The so-called \emph{i-}umlaut is very different in this respect. It is
clearly not alive as a phonological rule anymore but its effects can
still be observed in the modern language in certain morphologically
definable environments. As a result we can find near-minimal pairs of
word forms where \emph{i-}umlaut has applied in one member but not the
other although the phonological conditions seem identical. Some examples
are given in (8):

\ea
	\ea 	\emph{háttur} `mode', \textsc{dat.sg}
		\emph{hætt-i}/*\emph{hátt-i}, \textsc{nom.pl} \emph{hætt-ir/*hátt-ir}
	\ex	\emph{sáttur} `satisfied', \textsc{nom.sg.m.wk}
		\emph{*sætt-i/sátt-i}, \textsc{nom.pl.m} \emph{*sætt-ir/sátt-ir}
	\z
\z

\noindent In (8a) we see examples of the paradigmatic alternation /á
\textasciitilde{} æ/ (phonetically {[}au{]} \textasciitilde{} {[}ai{]}
in the modern language, probably {[}aː{]} \textasciitilde{} {[}ɛː{]} in
Old Icelandic) originally caused by \emph{i-}umlaut. In the
\textsc{nom.sg} we have /á/ in the stem but in the \textsc{dat.sg} the
only acceptable form is \emph{hætti} and the ``non-umlauted'' version
\emph{*hátti} is unacceptable. Similarly, in the \textsc{nom.pl} only
\emph{hættir} is acceptable and *\emph{háttir} is not. At a first glance
we might think that an /i/ in the inflectional ending is still causing
this ``umlaut'' but a comparison with the adjectival forms in (8b)
indicates that this cannot be the case. Here the only acceptable weak
\textsc{nom.sg.m} form is \emph{sátti} and not \emph{*sætti} and the
only \textsc{nom.pl.m} form is \emph{sáttir} and not *\emph{sættir}. So
the \emph{i-}umlaut alternations in Modern Icelandic are clearly
morphologically conditioned and not phonological anymore (see also
\citealt[93]{thrainsson2011} for further examples of this kind).

Second, recall that standard generative phonology formulations of
\emph{u-}umlaut in Icelandic of the kind illustrated in (3b) above state
explicitly that /u/ only triggers umlaut of /a/ in the immediately
preceding syllable. This is illustrated by examples like the following:

\ea
	\ea	\emph{bakki} `bank' \textsc{dat.pl}
		\emph{b\textbf{ö}kk-\textbf{u}m/*b\textbf{a}kk-\textbf{u}m}
	\ex	\emph{akkeri} `anchor' \textsc{dat.pl}
		*\emph{\textbf{ö}kker-\textbf{u}m/\textbf{a}kker-\textbf{u}m}
	\z
\z

\noindent In (9a) the \emph{u-}umlaut obligatorily applies to the root vowel /a/
in the immediately preceding syllable. In (9b), on the other hand, the
/u/ in the (same) inflectional ending cannot apply to the root vowel /a/
because there is a syllable intervening. An interesting and much
discussed case, e.g. by Anderson in several of the publications cited
above, involves trisyllabic words with two instances of /a/ in the stem.
Consider the examples in (10):

\ea
	\ea	\emph{k\textbf{a}ll\textbf{a}} ʽcallʼ\\
		\textsc{1.sg.pst} \emph{kalla-ð-i}, \textsc{1.pl.pst}
		\emph{*k\textbf{a}ll\textbf{ö}-ð-um/k\textbf{ö}ll\textbf{u}-ð-um/*k\textbf{a}ll\textbf{u}-ð-		um/*k\textbf{ö}ll\textbf{a}-ð-\textbf{u}m}
	\ex	\emph{b\textbf{a}n\textbf{a}n-i} `banana'\\
		\textsc{dat.pl} \emph{b\textbf{a}n\textbf{ö}n-um/b\textbf{ö}n\textbf{u}n-um/*b\textbf{a}n\textbf{u}n-um/*b\textbf{ö}n\textbf{a}n-\textbf{u}m}
	\z
\z

\noindent Consider first the conceivable \textsc{1.pl.pst} forms of the verb
\emph{kalla} `call'. Based on the formulation (3b) of the
\emph{u-}umlaut rule, one might have expected the form \emph{*kallöðum},
where the /u/ in the inflectional ending triggers \emph{u}-umlaut of the
/a/ in the preceding syllable. This is not an acceptable form, however.
The reason is that in forms of this sort a ``weakening'' of unstressed
/ö/ to /u/ is obligatory. This weakening is found in in many words, e.g.
the plural of the word \emph{hérað} `district', plural
\emph{hér\textbf{ö}ð} or (preferred) \emph{hér\textbf{u}ð, meðal}
`medicine', plural \emph{með\textbf{ö}l} or (preferred)
\emph{með\textbf{u}l}. It is not always obligatory but it seems that in
the past tense of verbs of this sort it is. But once the (umlauted) /ö/
in \emph{*kall\textbf{ö}ðum} has been weakened to /u/ it obligatorily
triggers \emph{u-}umlaut of the preceding /a/ so
\emph{k\textbf{ö}ll\textbf{u}ðum} is acceptable but
\emph{*k\textbf{a}ll\textbf{u}ðum} is not. Finally, the form
\emph{*k\textbf{ö}ll\textbf{a}ðum} is not acceptable either, since there
\emph{u-}umlaut would be applied across an intervening syllable, which
is not possible, as we have seen (cf. (9b)). The \emph{u-}umlaut works
in a similar fashion in the word \emph{banani}, except that here the
weakening of the second (and unstressed) syllable from /ö/ to /u/ is not
obligatory. Hence \emph{ban\textbf{ö}n\textbf{u}m} is an acceptable
form, with the /u/ in the final syllable triggering \emph{u-}umlaut of
the preceding /a/ to /ö/. But if this /ö/ is further weakened to /u/,
then \emph{u-}umlaut of the first /a/ is obligatory and
\emph{b\textbf{ö}n\textbf{u}num} is an acceptable form but
*\emph{b\textbf{a}n\textbf{u}num} is not.\footnote{It is sometimes
  claimed that \emph{b\textbf{ö}n\textbf{ö}n\textbf{u}m} is also an
  acceptable form for some speakers. If this is so, it is possible that
  the /ö/ in the next-to-last syllable triggers \emph{u-}umlaut (i.e.
  \emph{ö-}umlaut!) of the /a/ in the first syllable. That would simply
  mean that the feature {[}−low{]} in the definition of the environment
  of the \emph{u-}umlaut in (3b) would be omitted. But since there are
  no derivational (nor inflectional) morphemes containing an underlying
  /ö/ (i.e. an /ö/ that cannot have been derived by \emph{u-}umlaut),
  this proposal cannot be tested independently of the iterative rule
  application, as pointed out by a reviewer.} As predicted by the
formulation of the \emph{u-}umlaut rule in (3b) a form like
*\emph{b\textbf{ö}nanum} is unacceptable because there the
\emph{u-}umlaut would have applied across an intervening syllable. Facts
of this sort have been interpreted as showing that \emph{u-}umlaut in
Modern Icelandic is of a phonological nature since it depends on
syllabic structure (no syllables can intervene between the umlaut
trigger and the target) and it can be applied iteratively (a /u/ which
itself is derived by \emph{u-}umlaut and subsequent independently needed
weakening can trigger \emph{u-}umlaut).

\section{The conditions for \emph{u}-umlaut in Modern
Faroese}

\subsection{\emph{u-}umlaut and the Modern Faroese
vowel system}

Modern Faroese has preserved some \emph{u-}umlaut-like vowel
alternations. A couple of examples are given in (11) (see also
\citealt[78, 100, passim]{thrainsson2012}):

\begin{exe}
\ex 	\emph{dag-ur} `day', \textsc{dat.pl}
	\emph{d\textbf{ø}g-\textbf{u}m}; 
	\emph{spak-ur} ʽcalmʼ, \textsc{nom.pl.wk}
	\emph{sp\textbf{ø}k-\textbf{u}}
\z

\noindent At first glance, these alternation seem very similar to the Icelandic
ones described in the preceding sections. But while the \emph{u-}umlaut
alternations are arguably phonologically (or phonetically) natural in
Modern Icelandic (see the diagram in (2b) and the formulation in (3b)),
it will be claimed below that this is not the case in Faroese. To
demonstrate this, it is necessary to look closely at the Faroese vowel
system. Consider first the following schematic representation of Faroese
\emph{u-}umlaut of the type just illustrated, where the alleged umlaut
trigger is encircled (cf. \citealt[98]{thrainsson2011}, \citealt[33]{thrainsson2012}, compare \citealt[248--250]{arnason2011}):\footnote{Vowel length
  is predictable in Faroese, as it is in Icelandic: Vowels are long in
  stressed open syllables, otherwise short. As illustrated in the
  brackets in (12), there is often a considerable difference in the
  phonetic realization of the long and short variants. This will be
  illustrated below. --- It should be noted that \citet[76]{arnason2011}
  assumes a different analysis of Faroese monophthongs.}

\ea \emph{u-}umlaut in Modern Faroese and the system of monophthongs:
		\begin{tabular}[t]{ccccccc}
		&& \multicolumn{2}{c}{{[}-back{]}} &  & \multicolumn{2}{c}{{[}+back{]}} \\
		&& {[}-round{]}   	& {[}+round{]}   &  & {[}-round{]}   & {[}+round{]}   \\
		{[}+high{]}  &   &  i        &      y          &  &                &    \textcircled{u} [uː/ʊ]            \\
		&  &   e    &   \tikzmarkfullnamed{oe}{ø} {[}øː/œ{]} &  &                &            o    \\
		{[}+low{]}  	&  &  \tikzmarkfullnamed{ae}{æ} {[}ɛaː/a{]}   &        &  & a	    &          ɔ \\
	\end{tabular}\begin{tikzpicture}[remember picture,overlay]
\draw[thick,shorten >=3pt,->,>=stealth] ([yshift=2pt]ae.north) -- (oe.west);
\end{tikzpicture}
	\z 


Something like (13) would seem to be a possible formulation of a process
of this kind in traditional generative phonology terms (compare (3b)):

\ea Possible phonological formulation of \emph{u-}umlaut in Modern
Faroese:\\
	\phonr{/æ/}{\phonfeat{\feat{\textbf{+round}}\\\feat{\textbf{-low}}}}{C\sub{0}V\sub{\normalsize\phonfeat{\feat{\textbf{+round}}\\\feat{+back}\\\feat{\textbf{-low}}}}}
\z

Presented this way, \emph{u-}umlaut in Faroese looks like a fairly
natural assimilation rule at a first glance.\footnote{A reviewer
  suggests, however, that a process changing rounding and height as
  formulated for Faroese in (13), might be less natural from the point
  of view of acoustic phonetics than a process changing rounding and
  backness the way the \emph{u-}umlaut rule in Modern Icelandic does
  according to (3): The former affects both F1 (for the height
  difference) and F2 (for rounding) whereas the latter affects F2 in
  opposite directions (raising it for fronting but lowering it for
  rounding). Thus the Modern Icelandic \emph{u-}umlaut rule would
  ``generate more similar input-output mappings'', which may be
  preferred to less similar ones.} But the facts are somewhat more
complicated.

First, the alleged trigger /u/ is not too stable in Modern Faroese. The
reason is that unstressed /i,u/ are not distinguished in all Faroese
dialects. In some dialects they merge into an {[}ɪ{]}-like sound, in
others into an {[}ʊ{]}-like sound but some dialects distinguish them as
{[}ɪ{]} and {[}ʊ{]} (see \citealt[27]{thrainsson2012}, and references
cited there). This situation has clearly added to the phonological
opacity of \emph{u-}umlaut alternations for speakers acquiring Faroese.

Second, the target of the \emph{u-}umlaut in Faroese is arguably a
``moving'' one. As indicated in (12), the umlaut affects the phoneme
represented there as /æ/. As the orthography suggests, it is a
descendant of Old Norse /a/ in words like \emph{d\textbf{a}gur},
\emph{sp\textbf{a}kur} (see (11)). It is realized phonetically as
{[}ɛaː{]} when long and {[}a{]} when short, as shown in (12), cf.
\emph{spakur} {[}spɛaː\textsuperscript{h}kʊɹ{]} `calm', \textsc{sg.n}
\emph{spakt} {[}spakt{]} (see \citealt[34]{thrainsson2012} passim). But
in the history of Faroese Old Norse /a/ {[}a{]} and /æ/ {[}ɛː{]} merged
so the phoneme represented here as /æ/ can also be a descendant of Old
Norse /æ/ and then it is represented in the spelling as ʽæʼ, cf.
\emph{trælur} {[}t\textsuperscript{h}ɹɛaːlʊɹ{]} `slave', \emph{æða}
{[}ɛaːva{]} `eider duck'. Words written with ʽæʼ show the same
alternation between long {[}ɛaː{]} and short {[}a{]} as demonstrated for
\emph{spakur} and \emph{spakt} above (e.g. \emph{vænur} {[}vɛaːnʊɹ{]}
`beautiful' \textsc{sg.m} vs. \emph{vænt} {[}van̥t{]}, cf. Thráinsson et
al. 2012, p. 34). Yet it seems that \emph{u-}umlaut is rarely if ever
found in the ʽæʼ-words. Thus the \textsc{dat.pl} of \emph{trælur} is
\emph{tr\textbf{æ}l\textbf{u}m} and not \emph{*tr\textbf{ø}l\textbf{u}m}
(compare \textsc{dat.pl} \emph{d\textbf{ø}l\textbf{u}m} of \emph{dalur}
`valley') and although the words \emph{æða} `eider duck' and \emph{aða}
`(big) mussel' sound the same, i.e. as {[}ɛaːva{]}, the \textsc{dat.pl}
of the former has to be \emph{æðum} {[}ɛaːvʊn{]} and
\emph{\textbf{ø}vum} {[}øːvʊn{]} can only be \textsc{dat.pl} of
\emph{aða}.\footnote{A reviewer points out that the fact that
  \emph{u-}umlaut does not apply do `æ'-words in Faroese suggests that
  ``\emph{u}-umlaut had already taken on a morphological character
  before /a/ and /æ/ merged.'' But since there are no written records of
  Faroese from 1400‒1800, the historical development of the language is
  very murky.}

To further complicate matters, the development of Old Norse /a/ in
Faroese has left ``room'' for a ``regular /a/'' in the Faroese vowel
system, as shown in the diagram in (12). It occurs in loanwords and is
realized as {[}aː{]} when long and {[}a{]} when short, cf. \emph{Japan}
{[}ˈjaː\textsuperscript{h}pan{]}, \emph{japanskur}
{[}jaˈp\textsuperscript{h}anskʊɹ{]} `Japanese'.\footnote{In the noun
  \emph{Japan} the stress falls on the first syllable, in the adjective
  \emph{japanskur} it falls on the second one as indicated. Hence the
  quantity alternation in the first vowel.} It does not seem that this
vowel ever undergoes \emph{u-}umlaut in Faroese (for further discussion
see section 4).

Finally, there is no \emph{u}-epenthesis in Modern Faroese to ``explain
away'' apparent exceptions to \emph{u-}umlaut as will be shown in the
next section.

\subsection{The lack of \emph{u-}epenthesis in Modern Faroese}

Now recall that the most obvious surface exception to \emph{u-}umlaut in
Modern Icelandic is due to the \emph{u-}epenthesis described above. This
rule creates -\emph{ur-}endings that do not trigger \emph{u-}umlaut. It
was argued that this epenthesis rule is still productive in Icelandic,
witness the fact that it only applies in phonologically definable
environments. Hence there is a clear distributional difference between
-\emph{ur}-endings produced by the epenthesis rule (and not triggering
\emph{u-}umlaut) and -\emph{ur-}endings where the /u/ is a part of the
underlying form (and triggers umlaut). This is not the case in Faroese,
where the ending -\emph{ur} as a marker of the \textsc{nom.sg} of strong
masculine nouns and adjectives, with a /u/ that was historically
inserted by epenthesis, has been generalized to all environments. Hence
it has become distributionally indistinguishable from other
-\emph{ur-}endings. Table~\ref{tab:strong} compares the phonological realization of the
\textsc{nom.sg.m} \#-r\#-ending in Modern Icelandic to its Modern
Faroese counterpart (see also \citealt[100]{thrainsson2011}):

\begin{table}
	\begin{tabularx}{\textwidth}{@{}l>{\raggedright}p{1in}>{\raggedright}X>{\raggedright}X@{}}
		\lsptoprule
& type of stem & phonological realization of a strong \textsc{nom.sg.m} ending in Modern Icelandic & phonological
realization of a strong \textsc{nom.sg.m} ending in Modern
Faroese\tabularnewline
\midrule
a. & ending in a vowel & -\emph{r} \newline (\emph{mó-r} `peat', \emph{há-r}
`high') & \emph{-\textbf{ur}} \newline (\emph{mó-ur/mógv-ur} `peat', \emph{há-ur}
ʽhighʼ)\tabularnewline
b. & ending in a sin- gle consonant (but see c) & \emph{\textbf{-ur}} \newline
(\emph{dal-ur} `valley', \emph{lat-ur} ʽlazyʼ) & \emph{\textbf{-ur}} \newline
(\emph{dal-ur} `valley', \emph{lat-ur} ʽlazyʼ)\tabularnewline
c. & ending in a high vowel + /l,n/ & assimilation \newline (\emph{stól-l} `chair',
\emph{fín-n} `fine') & \textbf{-\emph{ur}} \newline (\emph{stól-ur} `chair',
\emph{fín-ur} `fine')\tabularnewline
d. & ending in /s, r/ or consonant clusters like /gl, gn/ & deletion \newline
(\emph{ís} `ice', \emph{laus} `loose', \emph{foss} `waterfall',
\emph{stór} `big', \emph{fugl} `bird', \emph{vagn} `wagon') &
\textbf{-\emph{ur}} \newline (\emph{ís-ur} `ice', \emph{leys-ur} `loose',
\emph{foss-ur} `waterfall', \emph{stór-ur} `big', \emph{fugl-ur} `bird',
\emph{vagn-ur} `wagon')\tabularnewline
	\lspbottomrule
	\end{tabularx}
\caption{Phonological realization of a strong \textsc{nom.sg.m}-ending in Modern Icelandic and Modern Faroese.}
\label{tab:strong}
\end{table}

This has clearly made the \emph{u-}umlaut rule in Faroese more opaque
since now the non-umlauting and umlauting \emph{ur-}endings occur in the
same phonological environments. It seems very likely that this has
contributed to the death of \emph{u-}umlaut as a phonological process in
Faroese.

\section{Testing the predictions}\label{testing-the-predictions}

In the preceding discussions I have described /a
\textasciitilde{} ö/ alternations in Modern Icelandic and their Modern
Faroese counterparts. I have argued that the Icelandic alternations are
still governed by a synchronic phonological process. Although these
alternations are still found in Modern Faroese, I have argued that they
cannot be governed by a phonological rule. Instead they must be
morphologically governed or analogical. This analysis makes several
testable predictions (see \citealt[100--102]{thrainsson2011}).

First, we do not a priori expect phonologically
conditioned alternations to be restricted to particular morphological
categories whereas morphologically conditioned alternations obviously
are, by definition. As we have already seen, the Icelandic
\emph{u-}umlaut occurs in the inflectional paradigms of nouns,
adjectives and verbs and in various grammatical categories (cases,
numbers, tenses, persons \ldots{}). Its Faroese counterpart behaves
differently. It is found in the inflectional paradigms of nouns and
adjectives, as we have seen (cf. (11)), but not in the past tense forms
of verbs, where it would be expected on phonological grounds. Thus we
have \emph{við kölluðum} in Icelandic vs. \emph{vit kallaðu} in Faroese
for \textsc{1.pl.pst} `we called', and \emph{við frömdum} vs. \emph{vit
framdu} in Faroese for \textsc{1.pl.pst} `we did,
made'.

Second, a phonological rule should not allow analogical extensions to
forms that do not fit its structural conditions. Such extensions are not
found for Icelandic \emph{u-}umlaut but in Faroese they are very common.
Thus the /ø/ of the oblique cases \emph{s\textbf{ø}g\textbf{u}} `saga'
has been analogically extended to the \textsc{nom.sg} form
\emph{s\textbf{ø}ga} and many other words of a similar type. The
corresponding form *\emph{s\textbf{ö}ga} is unacceptable in
Icelandic.\footnote{As a reviewer reminds me, the Icelandic neologism
  for \emph{computer} is interesting in this connection. It was supposed
  to be \emph{tölva} (related to the word \emph{tala} `number' --- this
  was when computers were mainly used for computing) in \textsc{nom.sg},
  oblique singular cases \emph{tölvu}. In Proto-Nordic time /v/ could
  trigger umlaut of /a/ to /ǫ/ so we have Old Norse words like
  \emph{vǫlva} `sooth-sayer, witch'. But since /v/ is not a trigger of
  umlaut in Modern Icelandic (witness loanwords like \emph{salvi}
  `salve, cream'), speakers tend to use the form \emph{talva} for
  \textsc{nom.sg}, thus in a way undoing the underlying /ö/ in the
  nominative as if they are ``assuming'' that the /ö/ in the oblique
  cases is derived by a synchronic \emph{u-}umlaut from /a/, as in words
  like \emph{saga} `saga', oblique \emph{sögu} (for some discussion see
\citealt{thrainsson1982}).}

Third, a phonologically conditioned rule should apply whenever its
structural conditions are met. Thus we would not expect to find
inflectional forms in Icelandic where \emph{u-}umlaut fails to apply in
an appropriate environment. Such examples are very common in Faroese, on
the other hand. Thus the \textsc{dat.pl} of the noun \emph{rakstur}
`shave' in Faroese is \emph{r\textbf{a}kstr\textbf{u}m} and not the
expected \emph{*r\textbf{ø}kstr\textbf{u}m}, the \textsc{dat.pl} of
\emph{spakur} `calm' can either be \emph{sp\textbf{ø}k\textbf{u}m} or
\emph{sp\textbf{a}k\textbf{u}m}, etc. (see \citealt[79, 100, passim]{thrainsson2012}). Corresponding unumlauted forms are unacceptable in
Icelandic.

Fourth, there is evidence for ``iterative'' application of
\emph{u-}umlaut in Icelandic, with one application of the
\emph{u-}umlaut rule feeding another. This was discussed above (second
part of section 2.2) in connection with forms like \textsc{1.pl.pst}
\emph{k\textbf{ö}ll\textbf{u}ð\textbf{u}m} `(we) called' and
\textsc{dat.pl} \emph{b\textbf{ö}n\textbf{u}n\textbf{u}m} `bananas'. No
such evidence is found in Faroese, where the corresponding forms are
\emph{k\textbf{a}ll\textbf{a}ðum} and
\emph{b\textbf{a}n\textbf{a}num}.\footnote{The latter form may be
  related to the fact that \emph{banan} `banana' is a loanword and
  contains the vowel /a/ (long variant {[}aː{]}) and not /æ/, cf. the
  discussion in section 3.1. See also the next paragraph.}

Finally, Icelandic \emph{u-}umlaut is so productive that it is naturally
applied in loanwords, as we have seen. This is not so in Faroese. Thus
the word \emph{app} (for a small program) has been adopted into both
languages. In Icelandic the \textsc{dat.pl} has to be
\emph{\textbf{ö}pp\textbf{u}m} whereas the natural form is
\emph{\textbf{a}pp\textbf{u}m} in Faroese. This can easily be verified
by searching for the word combinations \emph{með öppum} and \emph{við
appum} `with apps' on Google. For the first variant one finds a number
of Icelandic hits, for the second Faroese ones.

The general conclusion, then, is that \emph{u-}umlaut in Modern
Icelandic has a number of properties that are to be expected if it is a
phonological process but none of the properties one might expect of
morphologically conditioned or analogical alternations.


\section{Concluding remarks}\label{concluding-remarks}

While it has often been argued that phonology need not be ``natural''
(see e.g. \citealt{anderson1981}), there must obviously be limits to the
``unnaturalness'' and opacity of phonological processes. Once they
become too unnatural and opaque, they can no longer be acquired as such
and the phonological alternations originally created by them will be
relegated to morphology. Then their productivity will be limited and it
will at best survive to some extent by analogy, but analogical processes
are known to be irregular and unpredictable. The fate of \emph{i-}umlaut
in Icelandic is a case in point, as described above (see the discussion
of the examples in (8)). But whereas we do not have detailed information
about how \emph{i-}umlaut died as a phonological process, comparison of
the development of \emph{u-}umlaut in Icelandic and Faroese sheds an
interesting light on how a phonological rule can die and how it can
survive despite changing conditions.



\section*{Acknowledgements}
Many thanks to Steve Anderson for introducing me to the wonders of synchronic \emph{u-}umlaut way back when. Thanks are also due to the editors and to two anonymous reviewers for help, useful comments, suggestions and corrections.

\printbibliography[heading=subbibliography,notkeyword=this]


% \todos
\end{document}
